---
title: "08_analyze_ucsf_results.Rmd"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(table1)
library(ggpubr)
library(mgcViz)
library(ggrepel)
library(writexl)

```

Copyright (C) 2023 Jacquelyn Roger
License described at: https://github.com/jackieroger/RPL_association_study

### Load data

```{r}

load("data/cohorts/RPL.rda")
load("data/cohorts/C.rda")
load("data/cohorts/all_patients_rpl_control.rda")
load("data/umap/umap_coords.rda")
load("data/associations/stats_all.rda")
load("data/associations/stats_all_hu.rda")

```

### Log transforms

```{r}

stats_rpl_control <- stats_rpl_control %>%
  mutate(log_odds_ratio = log(odds_ratio_w_age_re, base = 10),
         log_p_value = -log(adj_p_value_w_age_re, base = 10))

stats_rpl_control <- stats_rpl_control %>%
  mutate(Result = factor(case_when(
    adj_p_value_w_age_re < 0.05 & log_odds_ratio > 0 ~ "Positive",
    adj_p_value_w_age_re < 0.05 & log_odds_ratio < 0 ~ "Negative",
    TRUE ~ "Not significant"), levels = c("Positive", "Negative", "Not significant")
  ))

stats_hu <- stats_hu %>%
  mutate(log_odds_ratio = log(odds_ratio_w_hu, base = 10),
         log_p_value = -log(adj_p_value_w_hu, base = 10)) %>%
  mutate(Result = factor(case_when(
    adj_p_value_w_hu < 0.05 & log_odds_ratio > 0 ~ "Positive",
    adj_p_value_w_hu < 0.05 & log_odds_ratio < 0 ~ "Negative",
    TRUE ~ "Not significant"), levels = c("Positive", "Negative", "Not significant")
  ))
stats_hu_sig <- filter(stats_hu, Result != "Not significant")

```

### UMAP

```{r}

# Pull together umap info
umap_coords <- umap_coords %>%
  left_join(all_patients_rpl_control) %>%
  mutate(`Age strata` = factor(ifelse(age_t0 < 35, "<35", "35+"), levels = c("<35", "35+")))

# Compute p-values for coords
wilcox.test(x ~ cohort, data = umap_coords, paired = FALSE, exact = FALSE, conf.int = TRUE)
wilcox.test(x ~ cohort, data = umap_coords, paired = FALSE, exact = FALSE, conf.int = TRUE)

# Groups
p1 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = Group)) +
  theme_minimal() +
  scale_color_manual(values=c("indianred2", "steelblue2")) +
  ggtitle("UCSF diagnosis UMAP - RPL & Control")

p2  <- ggplot(umap_coords) +
  geom_violin(aes(x = cohort, y = x, fill = Group, color = Group)) +
  geom_boxplot(aes(x = cohort, y = x, fill = Group), width = 0.1, fill = "white", color = "black", outlier.size = 0.5) +
  stat_compare_means(aes(x = cohort, y = x), label.x = 1.1, label.y = 8, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values=c("indianred2", "steelblue2")) +
  scale_color_manual(values=c("indianred2", "steelblue2")) +
  theme(axis.title.x = element_blank(),
        plot.margin = margin(t = 20))

p3 <- ggplot(umap_coords) +
  geom_violin(aes(x = cohort, y = y, fill = Group, color = Group)) +
  geom_boxplot(aes(x = cohort, y = y, fill = Group), width = 0.1, fill = "white", color = "black", outlier.size = 0.5) +
  stat_compare_means(aes(x = cohort, y = x), label.x = 1.1, label.y = 11, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values=c("indianred2", "steelblue2")) +
  scale_color_manual(values=c("indianred2", "steelblue2")) +
  theme(axis.title.x = element_blank(),
        plot.margin = margin(t = 20))
p4 <- ggarrange(p2, p3, nrow = 2)
p5 <- ggarrange(p1, p4, ncol = 2, widths = c(2,1))
p5
ggsave("figures/all/umap/umap_groups.png", width = 9, height = 4)

# Age
p6 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = `Age strata`)) +
  theme_minimal() +
  scale_color_manual(values = c("darkgoldenrod2", "mediumpurple2")) +
  ggtitle("UCSF diagnosis UMAP - age strata")

p7  <- ggplot(umap_coords) +
  geom_violin(aes(x = `Age strata`, y = x, fill = `Age strata`, color = `Age strata`)) +
  geom_boxplot(aes(x = `Age strata`, y = x, fill = `Age strata`), width = 0.1, fill = "white", color = "black", outlier.size = 0.5) +
  stat_compare_means(aes(x = `Age strata`, y = x), label.x = 1.1, label.y = 8, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values = c("darkgoldenrod2", "mediumpurple2")) +
  scale_color_manual(values = c("darkgoldenrod2", "mediumpurple2")) +
  theme(axis.title.x = element_blank(),
        plot.margin = margin(t = 20))

p8 <- ggplot(umap_coords) +
  geom_violin(aes(x = `Age strata`, y = y, fill = `Age strata`, color = `Age strata`)) +
  geom_boxplot(aes(x = `Age strata`, y = y, fill = `Age strata`), width = 0.1, fill = "white", color = "black", outlier.size = 0.5) +
  stat_compare_means(aes(x = `Age strata`, y = x), label.x = 1.1, label.y = 11, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values = c("darkgoldenrod2", "mediumpurple2")) +
  scale_color_manual(values = c("darkgoldenrod2", "mediumpurple2")) +
  theme(axis.title.x = element_blank(),
        plot.margin = margin(t = 20))
p9 <- ggarrange(p7, p8, nrow = 2)
p10 <- ggarrange(p6, p9, ncol = 2, widths = c(2,1))
p10
ggsave("figures/all/umap/umap_age.png", width = 9, height = 4)

# Combine group and age umaps
ggarrange(p5, p10, nrow = 2, widths = c(1,2))
ggsave("figures/all/umap/umap_both.png", width = 9, height = 8)
ggsave("figures/all/umap/pdf_fig3.pdf", width = 9, height = 9)

# Race
p1 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = Race)) +
  theme_minimal() +
  ggtitle("UCSF diagnosis UMAP - race")
p1
ggsave("figures/all/umap/umap_race.png", width = 8, height = 4)

# Ethnicity
p2 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = Ethnicity)) +
  theme_minimal() +
  scale_color_manual(values = c("violet", "yellowgreen", "cornflowerblue")) +
  ggtitle("UCSF diagnosis UMAP - ethnicity")
p2
ggsave("figures/all/umap/umap_ethnicity.png", width = 7, height = 4)

# Combine demographics umaps
ggarrange(p1, p2, nrow = 2)
ggsave("figures/all/umap/pdf_supp_fig2.pdf", width = 9, height = 9)

# Number of visits
p3 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = num_visits_study)) +
  labs(color = "Number of visits (in study)") +
  theme_minimal() +
  scale_color_gradient(low = "plum1", high = "forestgreen") +
  ggtitle("UCSF diagnosis UMAP - number of visits")
p3
ggsave("figures/all/umap/umap_num_visits.png", width = 7, height = 4)

# Years in EHR
p4 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = years_in_ehr_study)) +
  labs(color = "Years in EHR (in study)") +
  theme_minimal() +
  scale_color_gradient(low = "plum1", high = "forestgreen") +
  ggtitle("UCSF diagnosis UMAP - years in EHR")
p4
ggsave("figures/all/umap/umap_years_in_ehr.png", width = 7, height = 4)

# Number of diagnoses
p5 <- ggplot(umap_coords) +
  geom_point(aes(x = x, y = y, color = num_diagnoses_study)) +
  labs(color = "Number of diagnoses (in study)") +
  theme_minimal() +
  scale_color_gradient(low = "plum1", high = "forestgreen") +
  ggtitle("UCSF diagnosis UMAP - number of diagnoses")
p5
ggsave("figures/all/umap/umap_num_diagnoses.png", width = 7, height = 4)

# Combine HU umaps
ggarrange(p3, p4, p5, nrow = 3)
ggsave("figures/all/umap/pdf_supp_fig3.pdf", width = 9, height = 12)

```

### Basic stats about phenotypes

```{r}

# How many were tested?
length(unique(stats_rpl_control$phenotype))
# 1612

# How many categories?
length(unique(stats_rpl_control$category))
# 18

# How many were significant?
length(unique(filter(stats_rpl_control, adj_p_value_w_age_re < 0.05)$phenotype))
# 120

# How many were significantly positively associated?
length(which(stats_rpl_control$odds_ratio_w_age_re > 1 & stats_rpl_control$adj_p_value_w_age_re < 0.05))
# 51

# How many were significantly negatively associated?
length(which(stats_rpl_control$odds_ratio_w_age_re < 1 & stats_rpl_control$adj_p_value_w_age_re < 0.05))
# 69

```

### Manhattan and volcano plots

```{r}

p1 <- ggplot(stats_rpl_control, aes(x = category, y = log_p_value, color = category, label = phenotype)) +
  geom_jitter(show.legend = FALSE) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1), plot.margin = margin(t = 5, b = 5, l = 20, r = 5)) +
  xlab("Diagnosis category") +
  ylab("-log(p-value)") +
  ggtitle("All associations tested at UCSF")
p1
ggsave("figures/all/manhattan/manhattan.png", height = 3, width = 7)

stats_rpl_control_sig <- filter(stats_rpl_control, adj_p_value_w_age_re < 0.05)

p2 <- ggplot(stats_rpl_control_sig) +
  geom_point(aes(x = log_odds_ratio, y = log_p_value, color = Result)) +
  geom_label_repel(aes(x = log_odds_ratio, y = log_p_value, color = Result, label = ifelse(log_p_value > 100 | log_odds_ratio > 1.5 | (log_odds_ratio > 0.5 & log_p_value > 60), as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), size = 3) +
  theme_minimal() +
  scale_color_manual(values = c("indianred2", "steelblue2", "grey")) +
  labs(x = "Log odds ratio",
       y = "Negative log p-value") +
  ggtitle("Significant associations")
p2
ggsave("figures/all/volcano/volcano_sig_top.png", height = 5, width = 10)


```

### Numbers for age-stratified analysis

```{r}

# UNDER 35

# How many were tested?
length(unique(stats_rpl_control_u35$phenotype))
# 1525

# How many categories?
length(unique(stats_rpl_control_u35$category))
# 18

# How many were significant?
length(unique(filter(stats_rpl_control_u35, adj_p_value_w_age_re < 0.05)$phenotype))
# 94

# How many were significantly positively associated?
length(which(stats_rpl_control_u35$odds_ratio_w_age_re > 1 & stats_rpl_control_u35$adj_p_value_w_age_re < 0.05))
# 63

# How many were significantly negatively associated?
length(which(stats_rpl_control_u35$odds_ratio_w_age_re < 1 & stats_rpl_control_u35$adj_p_value_w_age_re < 0.05))
# 31

# OVER 35

# How many were tested?
length(unique(stats_rpl_control_o35$phenotype))
# 1506

# How many categories?
length(unique(stats_rpl_control_o35$category))
# 18

# How many were significant?
length(unique(filter(stats_rpl_control_o35, adj_p_value_w_age_re < 0.05)$phenotype))
# 93

# How many were significantly positively associated?
length(which(stats_rpl_control_o35$odds_ratio_w_age_re > 1 & stats_rpl_control_o35$adj_p_value_w_age_re < 0.05))
# 30

# How many were significantly negatively associated?
length(which(stats_rpl_control_o35$odds_ratio_w_age_re < 1 & stats_rpl_control_o35$adj_p_value_w_age_re < 0.05))
# 63

```

### Comparing age-stratified results

```{r}

# Combine stats
stats_age_stratified <- select(stats_rpl_control_u35, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
  dplyr::rename(OR_u35 = odds_ratio_w_age_re, P_u35 = adj_p_value_w_age_re) %>%
  inner_join(select(stats_rpl_control_o35, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re)) %>%
  dplyr::rename(OR_o35 = odds_ratio_w_age_re, P_o35 = adj_p_value_w_age_re) %>%
  mutate(logOR_u35 = log(OR_u35, base = 10),
         logOR_o35 = log(OR_o35, base = 10)) %>%
  mutate(Result = case_when(
    P_u35 < 0.05 & P_o35 < 0.05 ~ "Significant in both",
    P_u35 < 0.05 & P_o35 >= 0.05 ~ "Significant in <35",
    P_u35 >= 0.05 & P_o35 < 0.05 ~ "Significant in 35+",
    P_u35 >= 0.05 & P_o35 >= 0.05 ~ "Not significant"
  )) %>%
  left_join(select(stats_rpl_control, phenotype, rpl_count, control_count))
stats_age_stratified$Result <- factor(stats_age_stratified$Result, levels = c("Significant in both", "Significant in <35", "Significant in 35+", "Not significant"))

# How many are shared in both age strata analyses?
length(unique(stats_age_stratified$phenotype))
# 1419

# What are the results?
table(stats_age_stratified$Result)
# Significant in both  Significant in <35  Significant in 35+     Not significant 
#         59                  35                  34                1291

# How many <35 ORs are > 35+ ORs? (for sig results)
sum(filter(stats_age_stratified, Result != "Not significant")$OR_u35 > filter(stats_age_stratified, Result != "Not significant")$OR_o35)
# 111

# Correlations
cor.test(filter(stats_age_stratified, Result != "Not significant")$OR_u35, filter(stats_age_stratified, Result != "Not significant")$OR_o35, method = "spearman")
# 0.889203

cor.test(filter(stats_age_stratified, Result != "Not significant")$OR_u35, filter(stats_age_stratified, Result != "Not significant")$OR_o35, method = "pearson")
# -0.006594139
# p-value = 0.9411

```

### Loglog plot for age-stratified analysis

```{r}

# Significant - at least 10 in rpl or control
ggplot(filter(stats_age_stratified, Result != "Not significant" & (rpl_count >= 10 | control_count >= 10))) +
  geom_point(aes(logOR_u35, logOR_o35, color = Result)) +
  theme_minimal() +
  scale_color_manual(values = c("lightseagreen", "darkgoldenrod2", "mediumpurple2", "grey")) +
  xlim(-2,4) +
  ylim(-2,4) +
  geom_abline(slope = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_label_repel(aes(logOR_u35, logOR_o35, color = Result, label = ifelse(logOR_o35 - logOR_u35 > 0.25 | logOR_u35 < -1, as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = 1, size = 2) +
  geom_label_repel(aes(logOR_u35, logOR_o35, color = Result, label = ifelse(logOR_u35 - logOR_o35 > 0.85 | logOR_u35 > 1 | (logOR_u35 > 0 & logOR_o35 < 0 & Result == "Significant in both"), as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = -1.5, size = 2) +
  labs(x = "Log odds ratio for <35 patients",
       y = "Log odds ratio for 35+ patients") +
  ggtitle("Significant associations by age strata")
ggsave("figures/age_stratified/loglog/loglog_sig_10.png", height = 5, width = 7)

```

### Numbers for healthcare utilization sensitivity analysis

```{r}

# How many were tested?
length(unique(stats_hu$phenotype))
# 1612

# How many were significant?
length(unique(filter(stats_hu, adj_p_value_w_hu < 0.05)$phenotype))
# 138

# How many were significantly positively associated?
length(which(stats_hu$odds_ratio_w_hu > 1 & stats_hu$adj_p_value_w_hu < 0.05))
# 42

# How many were significantly negatively associated?
length(which(stats_hu$odds_ratio_w_hu < 1 & stats_hu$adj_p_value_w_hu < 0.05))
# 96

```

### Comparing utilization results

```{r}

# Wrangle data for loglog plot
stats_w_wo_hu <- select(stats_rpl_control, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
  dplyr::rename(OR_woHU = odds_ratio_w_age_re, P_woHU = adj_p_value_w_age_re) %>%
  inner_join(select(stats_hu, phenotype, odds_ratio_w_hu, adj_p_value_w_hu)) %>%
  dplyr::rename(OR_wHU = odds_ratio_w_hu, P_wHU = adj_p_value_w_hu) %>%
  mutate(logOR_woHU = log(OR_woHU, base = 10),
         logOR_wHU = log(OR_wHU, base = 10)) %>%
  mutate(Result = case_when(
    P_woHU < 0.05 & P_wHU < 0.05 ~ "Significant in both",
    P_woHU < 0.05 & P_wHU >= 0.05 ~ "Significant in main analysis",
    P_woHU >= 0.05 & P_wHU < 0.05 ~ "Significant in sensitivity analysis",
    P_woHU >= 0.05 & P_wHU >= 0.05 ~ "Not significant"
  )) %>%
  left_join(select(stats_rpl_control, phenotype, rpl_count, control_count))
stats_w_wo_hu$Result <- factor(stats_w_wo_hu$Result, levels = c("Significant in both", "Significant in main analysis", "Significant in sensitivity analysis", "Not significant"))

# How many are significant?
length(unique(filter(stats_w_wo_hu, Result != "Not significant")$phenotype))
# 148

# What are the results?
table(stats_w_wo_hu$Result)
# Significant in both   Significant in main analysis   Significant in sensitivity analysis    Not significant
#       110                          10                                  28                        1464 

# Correlations
cor.test(filter(stats_w_wo_hu, Result != "Not significant")$OR_woHU, filter(stats_w_wo_hu, Result != "Not significant")$OR_wHU, method = "spearman")
# 0.9974976

cor.test(filter(stats_w_wo_hu, Result != "Not significant")$OR_woHU, filter(stats_w_wo_hu, Result != "Not significant")$OR_wHU, method = "pearson")
# 0.9996783

# How many woHU ORs are > wHU ORs? (for sig results)
sum(filter(stats_w_wo_hu, Result != "Not significant")$OR_woHU > filter(stats_w_wo_hu, Result != "Not significant")$OR_wHU)
# 111

# What's the average percent decrease in OR after adjusting for HU?
median((stats_w_wo_hu$OR_woHU - stats_w_wo_hu$OR_wHU) / stats_w_wo_hu$OR_woHU) * 100
# 12.73718

```

### Utilization loglog

```{r}

# Significant (labeled) - at least 10 in rpl or control
ggplot(filter(stats_w_wo_hu, Result != "Not significant" & (rpl_count >= 10 | control_count >= 10))) +
  geom_point(aes(logOR_woHU, logOR_wHU, color = Result)) +
  theme_minimal() +
  scale_color_manual(values = c("darkgoldenrod4", "violet", "yellowgreen", "grey")) +
  xlim(-2,4) +
  ylim(-2,4) +
  geom_abline(slope = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_label_repel(aes(logOR_woHU, logOR_wHU, color = Result, label = ifelse(logOR_wHU - logOR_woHU > 0.1 | logOR_woHU < -0.9, as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = 1, size = 2) +
  geom_label_repel(aes(logOR_woHU, logOR_wHU, color = Result, label = ifelse(logOR_woHU - logOR_wHU > 0.1 | logOR_woHU > 2, as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = -2, size = 2) +
  labs(x = "Log odds ratio in main analysis",
       y = "Log odds ratio in sensitivity analysis") +
  ggtitle("Significant associations in main & sensitivity analysis")
ggsave("figures/healthcare_utilization/loglog_sig_top_10.png", height = 5, width = 8)

```

### Save results files

```{r}

stats_ucsf <- stats_rpl_control
stats_age_stratified_ucsf <- stats_age_stratified
stats_w_wo_hu_ucsf <- stats_w_wo_hu

save(stats_ucsf, stats_age_stratified_ucsf, stats_w_wo_hu_ucsf, file = "data/paper_figures_data/ucsf_stats.rda")

```



