---
title: "11_compare_centers.Rmd"
output: html_document
date: '2023-01-26'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggrepel)

```

Copyright (C) 2023 Jacquelyn Roger
License described at: https://github.com/jackieroger/RPL_association_study

### Load data

```{r}

load("data/stanford/stats_all.rda")
stats_stanford <- stats_rpl_control
stats_u35_stanford <- stats_rpl_control_u35
stats_o35_stanford <- stats_rpl_control_o35
load("data/stanford/stats_all_hu.rda")
stats_hu_stanford <- stats_hu

load("data/associations/stats_all.rda")
stats_ucsf <- stats_rpl_control
load("data/associations/stats_all_hu.rda")
stats_hu_ucsf <- stats_hu

```

### Pull together results from main analysis

```{r}

# All results
stats_ucsf_stanford <- select(stats_ucsf, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re, rpl_count, control_count) %>%
  dplyr::rename(OR_ucsf = odds_ratio_w_age_re, P_ucsf = adj_p_value_w_age_re, rpl_count_ucsf = rpl_count, control_count_ucsf = control_count) %>%
  inner_join(select(stats_stanford, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re, rpl_count, control_count)) %>%
  dplyr::rename(OR_stanford = odds_ratio_w_age_re, P_stanford = adj_p_value_w_age_re, rpl_count_stanford = rpl_count, control_count_stanford = control_count) %>%
  mutate(logOR_ucsf = log(OR_ucsf, base = 10),
         logOR_stanford = log(OR_stanford, base = 10)) %>%
  mutate(Result = case_when(
    P_ucsf < 0.05 & P_stanford < 0.05 ~ "Significant in both",
    P_ucsf < 0.05 & P_stanford >= 0.05 ~ "Significant at UCSF",
    P_ucsf >= 0.05 & P_stanford < 0.05 ~ "Significant at Stanford",
    P_ucsf >= 0.05 & P_stanford >= 0.05 ~ "Not significant"
  ))
stats_ucsf_stanford$Result <- factor(stats_ucsf_stanford$Result, levels = c("Significant in both", "Significant at UCSF", "Significant at Stanford", "Not significant"))

# Results that are significant in both centers
sig_in_both <- filter(stats_ucsf_stanford, Result == "Significant in both") %>%
  mutate(Result = case_when(
    OR_ucsf > 1 & OR_stanford > 1 ~ "Both significantly positive",
    OR_ucsf < 1 & OR_stanford < 1 ~ "Both significantly negative",
    TRUE ~ "Both significant, but discordant results",
  ))
sig_in_both$Result <- factor(sig_in_both$Result, levels = c("Both significantly positive", "Both significantly negative", "Both significant, but discordant results"))

```

### Hypergeometric tests
Resources:
1) https://rdrr.io/bioc/GeneOverlap/man/GeneOverlap.html
2) https://cran.r-project.org/web/packages/BiasedUrn/BiasedUrn.pdf
3) https://cran.r-project.org/web/packages/BiasedUrn/vignettes/UrnTheory.pdf

```{r}

# How many diagnoses were tested at both ucsf and stanford? (main)
length(intersect(unique(stats_ucsf$phecode), unique(stats_stanford$phecode)))
# 1576

# POSITIVE - MAIN

# How many diagnoses were significantly positive at both ucsf and stanford?
sum(sig_in_both$Result == "Both significantly positive")
# 42

# What's the maximum intersect number? (main) (note to self: the min() function here is not a typo)
min(length(which(stats_ucsf$odds_ratio_w_age_re > 1 & stats_ucsf$adj_p_value_w_age_re < 0.05)),
    length(which(stats_stanford$odds_ratio_w_age_re > 1 & stats_stanford$adj_p_value_w_age_re < 0.05)))
# 51

# Hypergeometric test for overlap between ucsf and stanford results (main analysis)
sum(dhyper(42:51, # number of white balls drawn
           330, # number of white balls in urn
           1576-330, # number of black balls in urn
           51)) # number of balls drawn
# p-value: 1.680169e-21

# NEGATIVE - MAIN

# How many diagnoses were significantly negative at both ucsf and stanford?
sum(sig_in_both$Result == "Both significantly negative")
# 34

# What's the maximum intersect number? (main) (note to self: the min() function here is not a typo)
min(length(which(stats_ucsf$odds_ratio_w_age_re < 1 & stats_ucsf$adj_p_value_w_age_re < 0.05)),
    length(which(stats_stanford$odds_ratio_w_age_re < 1 & stats_stanford$adj_p_value_w_age_re < 0.05)))
# 37

# Hypergeometric test for overlap between ucsf and stanford results (main analysis)
sum(dhyper(34:37, # number of white balls drawn
           69, # number of white balls in urn
           1576-69, # number of black balls in urn
           37)) # number of balls drawn
# p-value: 3.305351e-47

```

### Pull together results from utilization analysis

```{r}

stats_hu_ucsf_stanford <- select(stats_hu_ucsf, phenotype, odds_ratio_w_hu, adj_p_value_w_hu, rpl_count, control_count) %>%
  dplyr::rename(OR_ucsf = odds_ratio_w_hu, P_ucsf = adj_p_value_w_hu, rpl_count_ucsf = rpl_count, control_count_ucsf = control_count) %>%
  inner_join(select(stats_hu_stanford, phenotype, odds_ratio_w_hu, adj_p_value_w_hu, rpl_count, control_count)) %>%
  dplyr::rename(OR_stanford = odds_ratio_w_hu, P_stanford = adj_p_value_w_hu, rpl_count_stanford = rpl_count, control_count_stanford = control_count) %>%
  mutate(logOR_ucsf = log(OR_ucsf, base = 10),
         logOR_stanford = log(OR_stanford, base = 10)) %>%
  mutate(Result = case_when(
    P_ucsf < 0.05 & P_stanford < 0.05 ~ "Significant in both",
    P_ucsf < 0.05 & P_stanford >= 0.05 ~ "Significant at UCSF",
    P_ucsf >= 0.05 & P_stanford < 0.05 ~ "Significant at Stanford",
    P_ucsf >= 0.05 & P_stanford >= 0.05 ~ "Not significant"
  ))
stats_hu_ucsf_stanford$Result <- factor(stats_hu_ucsf_stanford$Result, levels = c("Significant in both", "Significant at UCSF", "Significant at Stanford", "Not significant"))

# Sig in both (intersection)
sig_in_both_hu <- filter(stats_hu_ucsf_stanford, Result == "Significant in both") %>%
  mutate(Result = case_when(
    OR_ucsf > 1 & OR_stanford > 1 ~ "Both significantly positive",
    OR_ucsf < 1 & OR_stanford < 1 ~ "Both significantly negative",
    TRUE ~ "Both significant, but discordant results",
  ))
sig_in_both_hu$Result <- factor(sig_in_both_hu$Result, levels = c("Both significantly positive", "Both significantly negative", "Both significant, but discordant results"))

```

### Comparing between centers after controlling for healthcare utilization

```{r}

# How many are significant with HU-adjustment?
length(unique(filter(stats_hu_ucsf_stanford, Result == "Significant in both")$phenotype))
# 90

# What are the results from the HU analysis?
table(sig_in_both_hu$Result)
# Both significantly positive    Both significantly negative       Both significant, but discordant results 
#           33                                       56                                        1

# What is the overlap with the results from the main analysis?
length(intersect(unique(sig_in_both$phenotype), unique(sig_in_both_hu$phenotype)))
# 70

# What is the overlap for validated positive associations?
length(intersect(unique(filter(sig_in_both, Result == "Both significantly positive")$phenotype), unique(filter(sig_in_both_hu, Result == "Both significantly positive")$phenotype)))
# 33

# What is the overlap for validated negative associations?
length(intersect(unique(filter(sig_in_both, Result == "Both significantly negative")$phenotype), unique(filter(sig_in_both_hu, Result == "Both significantly negative")$phenotype)))
# 34

# What is the overlap for discordant associations?
length(intersect(unique(filter(sig_in_both, Result == "Both significant, but discordant results")$phenotype), unique(filter(sig_in_both_hu, Result == "Both significant, but discordant results")$phenotype)))
# 1

# Which diagnoses are also in the overlap (are sig in both main & HU analysis, but different results label)
filter(sig_in_both_hu, !phenotype %in% c(
  intersect(unique(filter(sig_in_both, Result == "Both significantly positive")$phenotype), unique(filter(sig_in_both_hu, Result == "Both significantly positive")$phenotype)),
  intersect(unique(filter(sig_in_both, Result == "Both significantly negative")$phenotype), unique(filter(sig_in_both_hu, Result == "Both significantly negative")$phenotype)),
  intersect(unique(filter(sig_in_both, Result == "Both significant, but discordant results")$phenotype), unique(filter(sig_in_both_hu, Result == "Both significant, but discordant results")$phenotype))
) & phenotype %in% sig_in_both$phenotype)
# Constipation
# Pruritus and related conditions

```


### Correlations between sig results

```{r}

cor.test(sig_in_both$OR_ucsf, sig_in_both$OR_stanford, method = "spearman")
# 0.9458983

cor.test(sig_in_both$OR_ucsf, sig_in_both$OR_stanford, method = "pearson")
# 0.9999329

cor.test(sig_in_both_hu$OR_ucsf, sig_in_both_hu$OR_stanford, method = "spearman")
# 0.9443059

cor.test(sig_in_both_hu$OR_ucsf, sig_in_both_hu$OR_stanford, method = "pearson")
# 0.9999641

# How many are higher at stanford in main?
sum(sig_in_both$OR_stanford > sig_in_both$OR_ucsf)
# 73

# How many are higher at stanford in HU analysis?
sum(sig_in_both_hu$OR_stanford > sig_in_both_hu$OR_ucsf)
# 55

```

### Make loglog plot for main analysis

```{r}

ggplot(filter(sig_in_both, (rpl_count_ucsf >= 10 | control_count_ucsf >= 10) & (rpl_count_stanford >= 10 | control_count_stanford >= 10))) +
  geom_point(aes(logOR_ucsf, logOR_stanford, color = Result)) +
  theme_minimal() +
  scale_color_manual(values = c("palevioletred1", "skyblue2", "springgreen3")) +
  xlim(-2,4) +
  ylim(-2,4) +
  geom_abline(slope = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_label_repel(aes(logOR_ucsf, logOR_stanford, color = Result, label = ifelse(logOR_ucsf < -0.8 | (logOR_stanford - logOR_ucsf > 0.5), as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = 2, size = 2) +
  geom_label_repel(aes(logOR_ucsf, logOR_stanford, color = Result, label = ifelse(logOR_ucsf > 1 | (logOR_ucsf - logOR_stanford > 0.2), as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = -2, size = 2) +
  labs(x = "Log odds ratio in UCSF analysis",
       y = "Log odds ratio in Stanford analysis") +
  ggtitle("Both-significant associations in UCSF & Stanford analysis")
ggsave("figures/stanford/loglog_both_10.png", height = 5, width = 8)

```

### Loglog plot for utilization analysis

```{r}

ggplot(filter(sig_in_both_hu, (rpl_count_ucsf >= 10 | control_count_ucsf >= 10) & (rpl_count_stanford >= 10 | control_count_stanford >= 10))) +
  geom_point(aes(logOR_ucsf, logOR_stanford, color = Result)) +
  theme_minimal() +
  scale_color_manual(values = c("palevioletred1", "skyblue2", "springgreen3")) +
  xlim(-2,4) +
  ylim(-2,4) +
  geom_abline(slope = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_label_repel(aes(logOR_ucsf, logOR_stanford, color = Result, label = ifelse(logOR_ucsf < -0.8 | (logOR_stanford - logOR_ucsf > 0.3), as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = 2, size = 2) +
  geom_label_repel(aes(logOR_ucsf, logOR_stanford, color = Result, label = ifelse(logOR_ucsf > 1 | (logOR_ucsf - logOR_stanford > 0.3), as.character(phenotype), '')), show.legend = FALSE, min.segment.length = unit(0, 'lines'), nudge_y = -2, size = 2) +
  labs(x = "Log odds ratio in UCSF analysis",
       y = "Log odds ratio in Stanford analysis") +
  ggtitle("Both-significant associations in UCSF & Stanford analysis")
ggsave("figures/stanford/loglog_both_hu_10.png", height = 5, width = 8)

```

### Save results data

```{r}

save(stats_ucsf_stanford, stats_hu_ucsf_stanford, sig_in_both, sig_in_both_hu, file = "data/paper_figures_data/center_stats.rda")

```



