---
title: "12_make_supp_files.Rmd"
output: html_document
date: '2023-01-26'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(writexl)

```

Copyright (C) 2023 Jacquelyn Roger
License described at: https://github.com/jackieroger/RPL_association_study

### Load data

```{r}

load("data/cohort_selection/all_criteria.rda")
load("data/gender/TNG_concept_filtered.rda")

load("data/stanford/stats_all.rda")
stats_stanford <- stats_rpl_control
stats_u35_stanford <- stats_rpl_control_u35
stats_o35_stanford <- stats_rpl_control_o35
load("data/stanford/stats_all_hu.rda")
stats_hu_stanford <- stats_hu

load("data/associations/stats_all.rda")
stats_ucsf <- stats_rpl_control
stats_u35_ucsf <- stats_rpl_control_u35
stats_o35_ucsf <- stats_rpl_control_o35
load("data/associations/stats_all_hu.rda")
stats_hu_ucsf <- stats_hu

```

### Supplementary file 1: concept lists

```{r}

sheets <- list(
  "Recurrent pregnancy loss" = RPL_concept_filtered,
  "Pregnancy loss" = PL_SG_concept_filtered,
  "History of pregnancy loss" = HO_PL_concept_filtered,
  "Live-birth" = FTB_concept_filtered,
  "Preterm birth labor rupture" = PTB_concept_filtered,
  "Multiple gestation with loss" = PL_MG_concept_filtered,
  "Extrauterine or molar " = EP_concept_filtered,
  "Gender (trans, NB, or GD)" = TNG_concept_filtered)

write_xlsx(sheets, "submission/files/Supplementary_file_1.xlsx")

```

### Function for de-identifying results with diagnosis counts <10

```{r}

deid_results <- function(results_raw) {
  deid_results <- results_raw %>%
    mutate(count_under10 = ifelse(rpl_count < 10 | control_count < 10, TRUE, FALSE))
  for (r in 1:nrow(deid_results)) {
    if (deid_results$count_under10[r] == TRUE) {
      deid_results[r,4:(ncol(deid_results)-1)] <- NA
    }
  }
  deid_results$count_under10 <- factor(deid_results$count_under10, levels = c(FALSE,TRUE))
  deid_results <- arrange(deid_results, count_under10, phenotype)
  return (deid_results)
}

```

### Supplementary file 2: UCSF results

```{r}

stats_ucsf_deid <- deid_results(stats_ucsf)
stats_u35_ucsf_deid <- deid_results(stats_u35_ucsf)
stats_o35_ucsf_deid <- deid_results(stats_o35_ucsf)
stats_hu_ucsf_deid <- deid_results(stats_hu_ucsf)

sheets <- list(
  "UCSF main results" = stats_ucsf_deid,
  "UCSF <35 results" = stats_u35_ucsf_deid,
  "UCSF 35+ results" = stats_o35_ucsf_deid,
  "UCSF utilization results" = stats_hu_ucsf_deid)

write_xlsx(sheets, "submission/files/Supplementary_file_2.xlsx")

```

### Supplementary file 3: Stanford results

```{r}

stats_stanford_deid <- deid_results(stats_stanford)
stats_u35_stanford_deid <- deid_results(stats_u35_stanford)
stats_o35_stanford_deid <- deid_results(stats_o35_stanford)
stats_hu_stanford_deid <- deid_results(stats_hu_stanford)

sheets <- list(
  "Stanford main results" = stats_stanford_deid,
  "Stanford <35 results" = stats_u35_stanford_deid,
  "Stanford 35+ results" = stats_o35_stanford_deid,
  "Stanford utilization results" = stats_hu_stanford_deid)

write_xlsx(sheets, "submission/files/Supplementary_file_3.xlsx")

```



