---
title: "21_compare_age_stratified_coeffs.Rmd"
output: html_document
date: '2023-06-19'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(plyr)
library(dplyr)
library(stringr)
library(mgcv)
library(writexl)

```

Copyright (C) 2023 Jacquelyn Roger
License described at: https://github.com/jackieroger/RPL_association_study

### Functions for running association analysis

```{r, warning = FALSE}

# Function that computes coefficients and standard errors for each phecode
compute_coeff_se <- function(curr_phecode, hf, all_patients) {
  # Make a dataframe with outcome, age, and phecode status
  model_in <- select(all_patients, person_id, outcome, age, race, ethnicity) %>%
    # Outcome is either rpl (rpl vs control) or birth (birth vs loss)
    mutate(outcome = ifelse(outcome == "rpl" | outcome == "birth", 1, 0),
           phecode_status = ifelse(person_id %in% filter(hf, phecode == curr_phecode)$person_id, 1, 0))
  # Run model with age, race, ethnicity
  m_w_age_re <- mgcv::gam(outcome ~ phecode_status + s(age) + race + ethnicity, family = binomial(), method = "REML", data = model_in)
  # Get coeffs and ses for phecode status
  summ <- summary(m_w_age_re)$p.table[2,]
  curr_phecode_stats <- c(curr_phecode, summ[1], summ[2]) %>%
    t() %>%
    as.data.frame()
  colnames(curr_phecode_stats) <- c("phecode", "coeff", "se")
  # Return
  return (curr_phecode_stats)
}

# Function that gets coefficients and standard errors for all phecodes
get_coeffs_ses <- function(hf_case, hf_control, all_patients) {
  # Combine hf
  hf <- rbind(hf_case, hf_control) %>%
    distinct()
  # All phecodes in hf
  all_phecodes <- select(hf, phecode) %>% distinct()
  # Compute stats for first phecode
  coeffs_ses <- compute_coeff_se(all_phecodes$phecode[1], hf, all_patients)
  # Compute stats for the rest
  for (p in 2:nrow(all_phecodes)) {
    curr_phecode <- all_phecodes$phecode[p]
    p_stats <- compute_coeff_se(curr_phecode, hf, all_patients)
    coeffs_ses <- rbind(coeffs_ses, p_stats)
  }
  return (coeffs_ses)
}

# Function for doing analysis
compute_stats <- function(co_case, co_control, case_cohort_size, control_cohort_size, all_patients, outcome) {
  coeffs_ses <- get_coeffs_ses(co_case, co_control, all_patients)
  stats <- select(phecode_definitions, phecode, phenotype, category) %>%
    inner_join(coeffs_ses) %>%
    arrange(phenotype)
  return (stats)
}

```

### Load data and prep dataframes for analysis

```{r, message = FALSE, warning = FALSE}

# Load data
phecode_definitions <- read.csv("data/phecodes/definitions/phecode_definitions1.2.csv")
load("data/cohorts/RPL.rda")
load("data/cohorts/C.rda")
load("data/phecodes/mappings/co.rda")

```

### Pull everyone together and separate <35 and 35+ patients

```{r}

# Combine
all_patients_rpl_control <- mutate(RPL, outcome = "rpl") %>%
  rbind(mutate(C, outcome = "control")) %>%
  dplyr::rename(age = age_t0)

# Get unique occurrences of conditions
co_rpl <- select(co_rpl, person_id, phecode) %>%
  distinct()
co_control <- select(co_control, person_id, phecode) %>%
  distinct()

# Filter patients
all_patients_rpl_control_u35 <- all_patients_rpl_control %>%
  filter(age < 35)

# Filter conditions
co_rpl_u35 <- filter(co_rpl, person_id %in% all_patients_rpl_control_u35$person_id)
co_control_u35 <- filter(co_control, person_id %in% all_patients_rpl_control_u35$person_id)

# Filter patients
all_patients_rpl_control_o35 <- all_patients_rpl_control %>%
  filter(age >= 35)

# Filter conditions
co_rpl_o35 <- filter(co_rpl, person_id %in% all_patients_rpl_control_o35$person_id)
co_control_o35 <- filter(co_control, person_id %in% all_patients_rpl_control_o35$person_id)

```

### Run association analysis

```{r}

# Compute stats
coeffs_ses_u35 <- compute_stats(
  co_rpl_u35,
  co_control_u35,
  length(unique(filter(RPL, age_t0 < 35)$person_id)),
  length(unique(filter(C, age_t0 < 35)$person_id)),
  all_patients_rpl_control_u35,
  "rpl"
)

# Compute stats
coeffs_ses_o35 <- compute_stats(
  co_rpl_o35,
  co_control_o35,
  length(unique(filter(RPL, age_t0 >= 35)$person_id)),
  length(unique(filter(C, age_t0 >= 35)$person_id)),
  all_patients_rpl_control_o35,
  "rpl"
)

# Save
save(coeffs_ses_u35, coeffs_ses_o35, file = "data/associations/coeffs_ses_age.rda")

```

### Z test

```{r}

age_coeffs_ztest <- inner_join(coeffs_ses_u35 %>% dplyr::rename(u35_coeff = coeff, u35_se = se),
                               coeffs_ses_o35 %>% dplyr::rename(o35_coeff = coeff, o35_se = se)) %>%
  mutate(z_score = (u35_coeff - o35_coeff) / sqrt((u35_se^2) + (o35_se^2)),
         z_pval = 2 * pnorm(q = z_score, lower.tail = FALSE),
         z_pval_bh = p.adjust(z_pval, method = "BH"))

```

### Save results

```{r}

save(age_coeffs_ztest, file = "data/associations/age_coeffs_ztest.rda")

```

### Write excel file

```{r}

sheets <- list("z test results" = age_coeffs_ztest %>%
    select(phenotype, category, z_score, z_pval, z_pval_bh) %>%
    filter(z_pval_bh < 0.05) %>%
    arrange(category))

write_xlsx(sheets, "data/associations/age_coeffs_ztest_sig.xlsx")

```
