---
title: "24_compute_confidence_intervals.Rmd"
output: html_document
date: '2023-06-25'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(plyr)
library(dplyr)
library(stringr)
library(mgcv)
library(writexl)

```

Copyright (C) 2023 Jacquelyn Roger
License described at: https://github.com/jackieroger/RPL_association_study

### Functions for running association analysis

```{r, warning = FALSE}

# Function that computes coefficients and standard errors for each phecode
compute_coeff_se <- function(curr_phecode, hf, all_patients, hu) {
  # Make a dataframe with outcome, age, and phecode status
  model_in <- select(all_patients, person_id, outcome, age, race, ethnicity, num_visits) %>%
    # Outcome is either rpl (rpl vs control) or birth (birth vs loss)
    mutate(outcome = ifelse(outcome == "rpl" | outcome == "birth", 1, 0),
           phecode_status = ifelse(person_id %in% filter(hf, phecode == curr_phecode)$person_id, 1, 0))
  # Run model
  if (hu == FALSE) {
    # Run model with age, race, ethnicity
  m_w_age_re <- mgcv::gam(outcome ~ phecode_status + s(age) + race + ethnicity, family = binomial(), method = "REML", data = model_in)
  summ <- summary(m_w_age_re)$p.table[2,]
  } else if (hu == TRUE) {
    # Run model with age, race, ethnicity, num visits in study
    m_w_hu <- mgcv::gam(outcome ~ phecode_status + s(age) + race + ethnicity + num_visits, family = binomial(), method = "REML", data = model_in)
    summ <- summary(m_w_hu)$p.table[2,]
  }
  # Get coeffs and ses for phecode status
  curr_phecode_stats <- c(curr_phecode, summ[1], summ[2]) %>%
    t() %>%
    as.data.frame()
  colnames(curr_phecode_stats) <- c("phecode", "coeff", "se")
  # Return
  return (curr_phecode_stats)
}

# Function that gets coefficients and standard errors for all phecodes
get_coeffs_ses <- function(hf_case, hf_control, all_patients, hu) {
  # Combine hf
  hf <- rbind(hf_case, hf_control) %>%
    distinct()
  # All phecodes in hf
  all_phecodes <- select(hf, phecode) %>% distinct()
  # Compute stats for first phecode
  coeffs_ses <- compute_coeff_se(all_phecodes$phecode[1], hf, all_patients, hu)
  # Compute stats for the rest
  for (p in 2:nrow(all_phecodes)) {
    curr_phecode <- all_phecodes$phecode[p]
    p_stats <- compute_coeff_se(curr_phecode, hf, all_patients, hu)
    coeffs_ses <- rbind(coeffs_ses, p_stats)
  }
  return (coeffs_ses)
}

# Function for doing analysis
compute_stats <- function(co_case, co_control, case_cohort_size, control_cohort_size, all_patients, outcome, hu = FALSE) {
  coeffs_ses <- get_coeffs_ses(co_case, co_control, all_patients, hu)
  stats <- select(phecode_definitions, phecode, phenotype, category) %>%
    inner_join(coeffs_ses) %>%
    arrange(phenotype)
  return (stats)
}

```

### Load data and prep dataframes for analysis

```{r, message = FALSE, warning = FALSE}

# Load data
phecode_definitions <- read.csv("data/phecodes/definitions/phecode_definitions1.2.csv")
load("data/cohorts/RPL.rda")
load("data/cohorts/C.rda")
load("data/phecodes/mappings/co.rda")

```

### Pull everyone together

```{r}

# Combine
all_patients_rpl_control <- mutate(RPL, outcome = "rpl") %>%
  rbind(mutate(C, outcome = "control")) %>%
  dplyr::rename(age = age_t0)

# Get unique occurrences of conditions
co_rpl <- select(co_rpl, person_id, phecode) %>%
  distinct()
co_control <- select(co_control, person_id, phecode) %>%
  distinct()

```

### Run association analysis

```{r}

# Compute stats
coeffs_ses <- compute_stats(
  co_rpl,
  co_control,
  length(unique(RPL$person_id)),
  length(unique(C$person_id)),
  all_patients_rpl_control,
  "rpl"
)
# Save
save(coeffs_ses, file = "data/associations/coeffs_ses.rda")

```

### Function for computing 95% confidence intervals

```{r}

compute_confidence_intervals <- function(coefficients_standarderrors) {
  confidence_intervals <- coefficients_standarderrors %>%
  mutate(ci_lower = exp(coeff - (1.96 * se)),
         ci_upper = exp(coeff + (1.96 * se)))
  return (confidence_intervals)
}

```

### Compute confidence intervals for main analysis

```{r}

cis <- compute_confidence_intervals(coeffs_ses)

save(cis, file = "data/associations/cis.rda")

```

### Compute confidence intervals for age stratified analysis

```{r}

load("data/associations/coeffs_ses_age.rda")

cis_u35 <- compute_confidence_intervals(coeffs_ses_u35)
cis_o35 <- compute_confidence_intervals(coeffs_ses_o35)

save(cis_u35, cis_o35, file = "data/associations/cis_age.rda")

```

### Compute confidence intervals for healthcare utilization sensitivity analysis

```{r}

# Compute stats
coeffs_ses_hu <- compute_stats(
  co_rpl,
  co_control,
  length(unique(RPL$person_id)),
  length(unique(C$person_id)),
  all_patients_rpl_control,
  "rpl",
  hu = TRUE
)

# Save
save(coeffs_ses_hu, file = "data/associations/coeffs_ses_hu.rda")

# Confidence intervals
cis_hu <- compute_confidence_intervals(coeffs_ses_hu)
save(cis_hu, file = "data/associations/cis_hu.rda")

```


