---
title: "25_make_more_supp_files.Rmd"
output: html_document
date: '2023-07-08'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(dplyr)
library(writexl)

```

### First, determine which results have diagnosis counts < 10 using the previously reported results

```{r}

ucsf_main <- read_excel("submission/files/Supplementary_file_2.xlsx", sheet = 1)
ucsf_u35 <- read_excel("submission/files/Supplementary_file_2.xlsx", sheet = 2)
ucsf_o35 <- read_excel("submission/files/Supplementary_file_2.xlsx", sheet = 3)
ucsf_hu <- read_excel("submission/files/Supplementary_file_2.xlsx", sheet = 4)

stanford_main <- read_excel("submission/files/Supplementary_file_3.xlsx", sheet = 1)
stanford_u35 <- read_excel("submission/files/Supplementary_file_3.xlsx", sheet = 2)
stanford_o35 <- read_excel("submission/files/Supplementary_file_3.xlsx", sheet = 3)
stanford_hu <- read_excel("submission/files/Supplementary_file_3.xlsx", sheet = 4)

```

### Confidence intervals

```{r}

deid_ci_results <- function(ci_id) {
  ci_deid <- ci_id %>%
    mutate(ci_lower = ifelse(count_under10, NA, ci_lower),
           ci_upper = ifelse(count_under10, NA, ci_upper)) %>%
    arrange(count_under10, phenotype) %>%
    select(phecode, phenotype, category, ci_lower, ci_upper, count_under10)
}

load("data/associations/cis.rda")
load("data/associations/cis_age.rda")
load("data/associations/cis_hu.rda")
ucsf_cis <- deid_ci_results(left_join(cis, ucsf_main))
ucsf_cis_u35 <- deid_ci_results(left_join(cis_u35, ucsf_u35))
ucsf_cis_o35 <- deid_ci_results(left_join(cis_o35, ucsf_o35))
ucsf_cis_hu <- deid_ci_results(left_join(cis_hu, ucsf_hu))

sheets <- list(
  "UCSF main CIs" = ucsf_cis,
  "UCSF <35 CIs" = ucsf_cis_u35,
  "UCSF 35+ CIs" = ucsf_cis_o35,
  "UCSF utilization CIs" = ucsf_cis_hu)

write_xlsx(sheets, "submission/files/Supplementary_file_4.xlsx")

load("data/stanford/cis.rda")
load("data/stanford/cis_age.rda")
load("data/stanford/cis_hu.rda")
stanford_cis <- deid_ci_results(left_join(cis, stanford_main))
stanford_cis_u35 <- deid_ci_results(left_join(cis_u35, stanford_u35))
stanford_cis_o35 <- deid_ci_results(left_join(cis_o35, stanford_o35))
stanford_cis_hu <- deid_ci_results(left_join(cis_hu, stanford_hu))

sheets <- list(
  "Stanford main CIs" = stanford_cis,
  "Stanford <35 CIs" = stanford_cis_u35,
  "Stanford 35+ CIs" = stanford_cis_o35,
  "Stanford utilization CIs" = stanford_cis_hu)

write_xlsx(sheets, "submission/files/Supplementary_file_5.xlsx")

```

### Z test

```{r}

load("data/associations/age_coeffs_ztest.rda")
ucsf_ztest <- age_coeffs_ztest %>%
  select(phecode, phenotype, category, z_score, z_pval, z_pval_bh) %>%
  arrange(z_pval_bh)

write_xlsx(list("UCSF age ztest" = ucsf_ztest), "submission/files/Supplementary_file_6.xlsx")

load("data/stanford/age_coeffs_ztest.rda")
stanford_ztest <- age_coeffs_ztest %>%
    select(phecode, phenotype, category, z_score, z_pval, z_pval_bh) %>%
  arrange(z_pval_bh)

write_xlsx(list("Stanford age ztest" = stanford_ztest), "submission/files/Supplementary_file_7.xlsx")

```

### Function for de-identifying results with diagnosis counts <10

```{r}

deid_results <- function(results_raw) {
  deid_results <- results_raw %>%
    mutate(count_under10 = ifelse(rpl_count < 10 | control_count < 10, TRUE, FALSE))
  for (r in 1:nrow(deid_results)) {
    if (deid_results$count_under10[r] == TRUE) {
      deid_results[r,4:(ncol(deid_results)-1)] <- NA
    }
  }
  deid_results$count_under10 <- factor(deid_results$count_under10, levels = c(FALSE,TRUE))
  deid_results <- arrange(deid_results, count_under10, phenotype)
  return (deid_results)
}

```

### 10 year analysis

```{r}

load("data/associations/stats_10.rda")
ucsf_10 <- stats_10 %>%
  deid_results()

write_xlsx(list("UCSF 10 year results" = ucsf_10), "submission/files/Supplementary_file_8.xlsx")

load("data/stanford/stats_10.rda")
stanford_10 <- stats_10 %>%
  deid_results()

write_xlsx(list("Stanford 10 year results" = stanford_10), "submission/files/Supplementary_file_9.xlsx")

```

### Look at which associations were significant at both UCSF and stanford for the 10 year analysis

```{r}

# All results
sig_results_10_ucsf_stanford <- select(ucsf_10, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re, rpl_count, control_count) %>%
  dplyr::rename(OR_ucsf = odds_ratio_w_age_re, P_ucsf = adj_p_value_w_age_re, rpl_count_ucsf = rpl_count, control_count_ucsf = control_count) %>%
  inner_join(select(stanford_10, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re, rpl_count, control_count)) %>%
  dplyr::rename(OR_stanford = odds_ratio_w_age_re, P_stanford = adj_p_value_w_age_re, rpl_count_stanford = rpl_count, control_count_stanford = control_count) %>%
  filter(P_ucsf < 0.05 & P_stanford < 0.05)

```

### 3PL subgroup analysis

```{r}

load("data/associations/stats_3pl.rda")
ucsf_3pl <- stats_3pl %>%
  deid_results()

write_xlsx(list("UCSF 3+ PLs results" = ucsf_3pl), "submission/files/Supplementary_file_10.xlsx")

load("data/stanford/stats_3pl.rda")
stanford_3pl <- stats_3pl %>%
  deid_results()

write_xlsx(list("Stanford 3+ PLs results" = stanford_3pl), "submission/files/Supplementary_file_11.xlsx")

```

### Look at which associations were significant at both UCSF and stanford for the 3PL subgroup analysis

```{r}

# All results
sig_results_3pl_ucsf_stanford <- select(ucsf_3pl, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re, rpl_count, control_count) %>%
  dplyr::rename(OR_ucsf = odds_ratio_w_age_re, P_ucsf = adj_p_value_w_age_re, rpl_count_ucsf = rpl_count, control_count_ucsf = control_count) %>%
  inner_join(select(stanford_3pl, phenotype, odds_ratio_w_age_re, adj_p_value_w_age_re, rpl_count, control_count)) %>%
  dplyr::rename(OR_stanford = odds_ratio_w_age_re, P_stanford = adj_p_value_w_age_re, rpl_count_stanford = rpl_count, control_count_stanford = control_count) %>%
  filter(P_ucsf < 0.05 & P_stanford < 0.05)

```

