---
title: "29_add_outcomes_analysis.Rmd"
output: html_document
date: '2024-01-26'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(plyr)
library(dplyr)
library(stringr)
library(mgcv)
library(writexl)
library(tidyr)
library(table1)

```

### Functions for running association analysis

```{r, warning = FALSE}

# Function for computing concept counts for each cohort
compute_counts <- function(hf_cohort) {
  hf_cohort_df_name <- deparse(substitute(hf_cohort))
  cohort <- str_split(hf_cohort_df_name, "_")[[1]][2]
  cc <- select(hf_cohort, phecode, person_id) %>%
    distinct() %>%
    group_by(phecode) %>%
    dplyr::summarise(count = n())
  colnames(cc)[2] <- paste0(cohort, "_count")
  return (cc)
}

# Function for getting the concept counts for each cohort and combining by study
# for a given type of health factor
get_study_counts <- function(hf_case, hf_control, case_cohort_size, control_cohort_size) {
  # Get concept counts for each cohort
  cc_case <- compute_counts(hf_case)
  cc_control <- compute_counts(hf_control)
  # Combine by study
  counts_case_control <- join_all(list(cc_case, cc_control), by = "phecode", type = "full") %>%
    arrange(phecode)
  counts_case_control[is.na(counts_case_control)] <- 0
  counts_case_control <- counts_case_control
  # Compute proportions
  counts_case_control <- counts_case_control %>%
    mutate(prop_case = case_count / case_cohort_size,
           prop_control = control_count / control_cohort_size)
  # Remove phenotypes where neither the case nor the control has >=5 people with it
  # counts_case_control <- filter(counts_case_control, case_count >= 5 | control_count >= 5)
  # Return counts
  return (counts_case_control)
}

# Function that computes odds ratio for each phecode
compute_ors_pvals <- function(curr_phecode, hf, all_patients) {
  # Make a dataframe with outcome, age, and phecode status
  model_in <- select(all_patients, person_id, outcome, age, race, ethnicity) %>%
    # Outcome is loss
    mutate(outcome = ifelse(outcome == "loss", 1, 0),
           phecode_status = ifelse(person_id %in% filter(hf, phecode == curr_phecode)$person_id, 1, 0))
  # Run crude model
  m_crude <- glm(outcome ~ phecode_status, family = binomial(), data = model_in)
  # Run model with age, race, ethnicity
  m_w_age_re <- mgcv::gam(outcome ~ phecode_status + s(age) + race + ethnicity, family = binomial(), method = "REML", data = model_in)
  # Get ORs and pvals
  curr_phecode_stats <- c(curr_phecode, exp(coef(m_crude))[2], coef(summary(m_crude))[,4][2], exp(coef(m_w_age_re))[2], summary(m_w_age_re)$p.table[2,4]) %>%
    t() %>%
    as.data.frame()
  colnames(curr_phecode_stats) <- c("phecode", "odds_ratio_crude", "p_value_crude", "odds_ratio_w_age_re", "p_value_w_age_re")
  # Return
  return (curr_phecode_stats)
}

# Function that gets odds ratios and pvals for all phecodes
get_ors_pvals <- function(counts, hf_case, hf_control, all_patients) {
  # Combine hf
  hf <- rbind(hf_case, hf_control) %>%
    distinct()
  # Compute stats for first phecode
  ors_pvals <- compute_ors_pvals(counts$phecode[1], hf, all_patients)
  # Compute stats for the rest
  for (p in 2:nrow(counts)) {
    curr_phecode <- counts$phecode[p]
    p_stats <- compute_ors_pvals(curr_phecode, hf, all_patients)
    ors_pvals <- rbind(ors_pvals, p_stats)
  }
  # Add a column for adjusted pvalues
  ors_pvals <- ors_pvals %>%
    mutate(adj_p_value_crude = p.adjust(p_value_crude, method = "BH"),
           adj_p_value_w_age_re = p.adjust(p_value_w_age_re, method = "BH")) %>%
    select(1, 2, 3, 6, 4, 5, 7)
  return (ors_pvals)
}

# Function for doing analysis
compute_stats <- function(co_case, co_control, case_cohort_size, control_cohort_size, all_patients, outcome) {
  counts_case_control <- get_study_counts(co_case, co_control, case_cohort_size, control_cohort_size)
  odds_ratios_pvals <- get_ors_pvals(counts_case_control, co_case, co_control, all_patients)
  stats <- select(phecode_definitions, phecode, phenotype, category) %>%
    inner_join(counts_case_control) %>%
    left_join(odds_ratios_pvals) %>%
    arrange(phenotype)
  # Customize column names
  stats <- stats %>%
      dplyr::rename(loss_count = case_count, birth_count = control_count, loss_prop = prop_case, birth_prop = prop_control)
  return (stats)
}

```

### Load data

```{r, message = FALSE, warning = FALSE}

# Load data
RPL_condition_occurrence <- read_tsv("data/from_rae/health_histories/RPL_condition_occurrence.txt")
load("data/outcomes_analysis/next_pregnancy_outcomes.rda")
load("data/phecodes/mappings/icd_to_phecode_mappings.rda")
phecode_definitions <- read.csv("data/phecodes/definitions/phecode_definitions1.2.csv")
load("data/cohorts/RPL.rda")
load("data/paper_figures_data/center_stats.rda")

```

### Find diagnoses that were significant in the main analysis

```{r}

load("data/stanford/stats_all.rda")
sig_in_stanford <- stats_rpl_control %>%
  select(phecode, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
  dplyr::rename(stanford_odds_ratio = odds_ratio_w_age_re,
                stanford_p_value = adj_p_value_w_age_re) %>%
  filter(stanford_p_value < 0.05)

load("data/associations/stats_all.rda")
sig_in_ucsf <- stats_rpl_control %>%
  select(phecode, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
  dplyr::rename(ucsf_odds_ratio = odds_ratio_w_age_re,
                ucsf_p_value = adj_p_value_w_age_re) %>%
  filter(ucsf_p_value < 0.05)

sig_in_both <- inner_join(sig_in_ucsf, sig_in_stanford)

```

### Make dataframes of the birth and loss subgroups of the RPL cohort

```{r}

# Assign age at next pregnancy outcome
next_pregs <- next_pregnancy_outcomes %>%
  left_join(select(RPL, person_id, birthdate, race, ethnicity)) %>%
  mutate(age = ifelse(outcome == "birth", as.numeric(livebirth_date - birthdate)/365, as.numeric(loss_date - birthdate)/365))

# Split outcomes
RPL_birth <- filter(next_pregs, outcome == "birth")
RPL_loss <- filter(next_pregs, outcome == "loss")

# Save
save(RPL_birth, RPL_loss, file = "data/outcomes_analysis/RPL_outcomes.rda")

```

### Aggregate conditions for RPL patients that occur before their birth or loss
Filter to only include diagnoses that were significant from the main analysis

```{r}

# RPL birth
co_RPL_birth <- left_join(RPL_condition_occurrence, icd_to_phecode_mappings) %>%
  drop_na() %>%
  inner_join(RPL_birth) %>%
  filter(condition_start_date < livebirth_date) %>%
  filter(phecode %in% sig_in_both$phecode)

# RPL loss
co_RPL_loss <- left_join(RPL_condition_occurrence, icd_to_phecode_mappings) %>%
  drop_na() %>%
  inner_join(RPL_loss) %>%
  filter(condition_start_date < loss_date) %>%
  filter(phecode %in% sig_in_both$phecode)

# Save
save(co_RPL_birth, co_RPL_loss, file = "data/phecodes/mappings/co_RPL_outcomes.rda")

```

### Run birth vs loss analysis

```{r}

# Get unique occurrences of conditions
co_RPL_birth <- select(co_RPL_birth, person_id, phecode) %>%
  distinct()
co_RPL_loss <- select(co_RPL_loss, person_id, phecode) %>%
  distinct()

# Compute stats
stats_RPL_outcomes <- compute_stats(
  co_RPL_loss,
  co_RPL_birth,
  length(unique(RPL_loss$person_id)),
  length(unique(RPL_birth$person_id)),
  next_pregs,
  "loss"
)

```

### Save results

```{r}

save(stats_RPL_outcomes, file = "data/outcomes_analysis/stats_RPL_outcomes.rda")

```

### Write excel file of sig results (adjusted p < 0.05)

```{r}

sheets <- list(
  "loss neg" = filter(stats_RPL_outcomes, odds_ratio_w_age_re < 1 & adj_p_value_w_age_re < 0.05) %>%
    select(phenotype, category, loss_count, birth_count, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category),
  "loss pos" = filter(stats_RPL_outcomes, odds_ratio_w_age_re > 1 & adj_p_value_w_age_re < 0.05) %>%
    select(phenotype, category, loss_count, birth_count, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category))

write_xlsx(sheets, "data/outcomes_analysis/sig_results_outcomes.xlsx")

```

### Write excel file of all results, organized by what the result was in the main analysis

```{r}

results_combined <- stats_RPL_outcomes %>%
  left_join(sig_in_both)

sheets <- list(
  "pos + pos" = filter(results_combined, ucsf_odds_ratio > 1 & odds_ratio_w_age_re > 1) %>%
    select(phenotype, category, ucsf_odds_ratio, loss_count, birth_count, odds_ratio_w_age_re, p_value_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category),
  "pos + neg" = filter(results_combined, ucsf_odds_ratio > 1 & odds_ratio_w_age_re <= 1) %>%
    select(phenotype, category, ucsf_odds_ratio, loss_count, birth_count, odds_ratio_w_age_re, p_value_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category),
  "neg + neg" = filter(results_combined, ucsf_odds_ratio <= 1 & odds_ratio_w_age_re <= 1) %>%
    select(phenotype, category, ucsf_odds_ratio, loss_count, birth_count, odds_ratio_w_age_re,  p_value_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category),
  "neg + pos" = filter(results_combined, ucsf_odds_ratio <= 1 & odds_ratio_w_age_re > 1) %>%
    select(phenotype, category, ucsf_odds_ratio, loss_count, birth_count, odds_ratio_w_age_re,  p_value_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category))

write_xlsx(sheets, "data/outcomes_analysis/all_results_outcomes.xlsx")

```

### Make a function for computing p-values for table1
Source 1: https://github.com/benjaminrich/table1/issues/52
Source 2: https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html

```{r}

# Function for computing p-values for table1
# Source 1: https://github.com/benjaminrich/table1/issues/52
# Source 2: https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html
pvalue <- function(x, ...) {
  x <- x[-length(x)]  # Remove "overall" group
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) {
    # For numeric variables, perform a 2-sample t-test
    p <- t.test(y ~ g)$p.value
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  # Format the p-value, using an HTML entity for the less-than sign.
  # The initial empty string places the output on the line below the variable label.
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

```

### Make demographic tables

```{r}

# Make table
table1(~ age + race + ethnicity | outcome, data = next_pregs, overall = "Total", extra.col = list(`P-value` = pvalue), extra.col.pos = 4)

```

### Compute spearman correlation of next pregnancy results and main analysis at UCSF

```{r}

cor.test(results_combined$odds_ratio_w_age_re, results_combined$ucsf_odds_ratio, method = "spearman")
# 0.7851571

```


