---
title: "30_make_outcomes_supp_file.Rmd"
output: html_document
date: '2024-02-23'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(writexl)

```

### Load data

```{r}

load("data/outcomes_analysis/stats_RPL_outcomes.rda")

```

### How many were significantly positively or negatively associated with loss?

```{r}

stats_RPL_outcomes %>% filter(adj_p_value_w_age_re < 0.05 & odds_ratio_w_age_re > 1) %>% nrow()

stats_RPL_outcomes %>% filter(adj_p_value_w_age_re < 0.05 & odds_ratio_w_age_re < 1) %>% nrow()

```

### Function for de-identifying results with diagnosis counts <10

```{r}

deid_results <- function(results_raw) {
  deid_results <- results_raw %>%
    mutate(count_under10 = ifelse(loss_count < 10 | birth_count < 10, TRUE, FALSE))
  for (r in 1:nrow(deid_results)) {
    if (deid_results$count_under10[r] == TRUE) {
      deid_results[r,4:(ncol(deid_results)-1)] <- NA
    }
  }
  deid_results$count_under10 <- factor(deid_results$count_under10, levels = c(FALSE,TRUE))
  deid_results <- arrange(deid_results, count_under10, phenotype)
  return (deid_results)
}

```

### Supplementary file 12: outcomes analysis

```{r}

stats_RPL_outcomes_deid <- deid_results(stats_RPL_outcomes)

sheets <- list(
  "outcomes analysis" = stats_RPL_outcomes_deid)

write_xlsx(sheets, "submission/files/Supplementary_file_12.xlsx")

```

