---
title: '31_add_t0_analysis.Rmd'
output: html_document
date: '2024-06-16'
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(plyr)
library(dplyr)
library(stringr)
library(mgcv)
library(writexl)
library(tidyr)

```

### Functions for running association analysis

```{r, warning = FALSE}

# Function for computing concept counts for each cohort
compute_counts <- function(hf_cohort) {
  hf_cohort_df_name <- deparse(substitute(hf_cohort))
  cohort <- str_split(hf_cohort_df_name, "_")[[1]][2]
  cc <- select(hf_cohort, phecode, person_id) %>%
    distinct() %>%
    group_by(phecode) %>%
    dplyr::summarise(count = n())
  colnames(cc)[2] <- paste0(cohort, "_count")
  return (cc)
}

# Function for getting the concept counts for each cohort and combining by study
# for a given type of health factor
get_study_counts <- function(hf_case, hf_control, case_cohort_size, control_cohort_size) {
  # Get concept counts for each cohort
  cc_case <- compute_counts(hf_case)
  cc_control <- compute_counts(hf_control)
  # Combine by study
  counts_case_control <- join_all(list(cc_case, cc_control), by = "phecode", type = "full") %>%
    arrange(phecode)
  counts_case_control[is.na(counts_case_control)] <- 0
  counts_case_control <- counts_case_control
  # Compute proportions
  counts_case_control <- counts_case_control %>%
    mutate(prop_case = case_count / case_cohort_size,
           prop_control = control_count / control_cohort_size)
  # Remove phenotypes where neither the case nor the control has >=5 people with it
  # counts_case_control <- filter(counts_case_control, case_count >= 5 | control_count >= 5)
  # Return counts
  return (counts_case_control)
}

# Function that computes odds ratio for each phecode
compute_ors_pvals <- function(curr_phecode, hf, all_patients) {
  # Make a dataframe with outcome, age, and phecode status
  model_in <- select(all_patients, person_id, outcome, age, race, ethnicity) %>%
    # Outcome is either rpl (rpl vs control) or birth (birth vs loss)
    mutate(outcome = ifelse(outcome == "rpl" | outcome == "birth", 1, 0),
           phecode_status = ifelse(person_id %in% filter(hf, phecode == curr_phecode)$person_id, 1, 0))
  # Run crude model
  m_crude <- glm(outcome ~ phecode_status, family = binomial(), data = model_in)
  # Run model with age, race, ethnicity
  m_w_age_re <- mgcv::gam(outcome ~ phecode_status + s(age) + race + ethnicity, family = binomial(), method = "REML", data = model_in)
  # Get ORs and pvals
  curr_phecode_stats <- c(curr_phecode, exp(coef(m_crude))[2], coef(summary(m_crude))[,4][2], exp(coef(m_w_age_re))[2], summary(m_w_age_re)$p.table[2,4]) %>%
    t() %>%
    as.data.frame()
  colnames(curr_phecode_stats) <- c("phecode", "odds_ratio_crude", "p_value_crude", "odds_ratio_w_age_re", "p_value_w_age_re")
  # Return
  return (curr_phecode_stats)
}

# Function that gets odds ratios and pvals for all phecodes
get_ors_pvals <- function(counts, hf_case, hf_control, all_patients) {
  # Combine hf
  hf <- rbind(hf_case, hf_control) %>%
    distinct()
  # Compute stats for first phecode
  ors_pvals <- compute_ors_pvals(counts$phecode[1], hf, all_patients)
  # Compute stats for the rest
  for (p in 2:nrow(counts)) {
    curr_phecode <- counts$phecode[p]
    p_stats <- compute_ors_pvals(curr_phecode, hf, all_patients)
    ors_pvals <- rbind(ors_pvals, p_stats)
  }
  # Add a column for adjusted pvalues
  ors_pvals <- ors_pvals %>%
    mutate(adj_p_value_crude = p.adjust(p_value_crude, method = "BH"),
           adj_p_value_w_age_re = p.adjust(p_value_w_age_re, method = "BH")) %>%
    select(1, 2, 3, 6, 4, 5, 7)
  return (ors_pvals)
}

# Function for doing analysis
compute_stats <- function(co_case, co_control, case_cohort_size, control_cohort_size, all_patients, outcome) {
  counts_case_control <- get_study_counts(co_case, co_control, case_cohort_size, control_cohort_size)
  odds_ratios_pvals <- get_ors_pvals(counts_case_control, co_case, co_control, all_patients)
  stats <- select(phecode_definitions, phecode, phenotype, category) %>%
    inner_join(counts_case_control) %>%
    left_join(odds_ratios_pvals) %>%
    arrange(phenotype)
  # Customize column names
  if (outcome == "birth") {
    stats <- stats %>%
      dplyr::rename(birth_count = case_count, loss_count = control_count, birth_prop = prop_case, loss_prop = prop_control)
  }
  if (outcome == "rpl") {
    stats <- stats %>%
      dplyr::rename(rpl_count = case_count, control_count = control_count, rpl_prop = prop_case, control_prop = prop_control)
  }
  return (stats)
}

```

### Load data

```{r, message = FALSE, warning = FALSE}

# Load data
load("data/cohorts/RPL.rda")
load("data/cohorts/C.rda")
load("data/phecodes/mappings/co.rda")
phecode_definitions <- read.csv("data/phecodes/definitions/phecode_definitions1.2.csv")

```

### Aggregate conditions for RPL and control patients before t0

```{r}

# RPL
co_rpl <- filter(co_rpl, condition_start_date < t0)

# RPL
co_control <- filter(co_control, condition_start_date < t0)

```

### Pull everyone together

```{r}

all_patients_rpl_control <- mutate(select(RPL, person_id, age_t0, race, ethnicity), outcome = "rpl") %>%
  rbind(mutate(select(C, person_id, age_t0, race, ethnicity), outcome = "control")) %>%
  dplyr::rename(age = age_t0)

```

### Run RPL vs control analysis

```{r}

# Get unique occurrences of conditions
co_rpl <- select(co_rpl, person_id, phecode) %>%
  distinct()
co_control <- select(co_control, person_id, phecode) %>%
  distinct()

# Compute stats
stats_t0 <- compute_stats(
  co_rpl,
  co_control,
  length(unique(RPL$person_id)),
  length(unique(C$person_id)),
  all_patients_rpl_control,
  "rpl"
)

```

### Save results

```{r}

save(stats_t0, file = "data/associations/stats_t0.rda")

```

### Write excel file of sig results (adjusted p < 0.05)

```{r}

sheets <- list(
  "RPL neg" = filter(stats_t0, odds_ratio_w_age_re < 1 & adj_p_value_w_age_re < 0.05) %>%
    select(phenotype, category, rpl_count, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category),
  "RPL pos" = filter(stats_t0, odds_ratio_w_age_re > 1 & adj_p_value_w_age_re < 0.05) %>%
    select(phenotype, category, rpl_count, odds_ratio_w_age_re, adj_p_value_w_age_re) %>%
    arrange(category))

write_xlsx(sheets, "data/associations/sig_results_t0.xlsx")

```

### How many were significantly positively or negatively associated with RPL?

```{r}

stats_t0 %>% filter(adj_p_value_w_age_re < 0.05 & odds_ratio_w_age_re > 1) %>% nrow()
# 55

stats_t0 %>% filter(adj_p_value_w_age_re < 0.05 & odds_ratio_w_age_re < 1) %>% nrow()
# 41

```

### Function for de-identifying results with diagnosis counts <10

```{r}

deid_results <- function(results_raw) {
  deid_results <- results_raw %>%
    mutate(count_under10 = ifelse(rpl_count < 10 | control_count < 10, TRUE, FALSE))
  for (r in 1:nrow(deid_results)) {
    if (deid_results$count_under10[r] == TRUE) {
      deid_results[r,4:(ncol(deid_results)-1)] <- NA
    }
  }
  deid_results$count_under10 <- factor(deid_results$count_under10, levels = c(FALSE,TRUE))
  deid_results <- arrange(deid_results, count_under10, phenotype)
  return (deid_results)
}

```

### Supplementary file 13: different temporal cutoff sensitiviy analysis

```{r}

stats_t0_deid <- deid_results(stats_t0)

sheets <- list(
  "temporal analysis" = stats_t0_deid)

write_xlsx(sheets, "submission/files/Supplementary_file_13.xlsx")

```

